
# Making good figures

```{r}
#| label: setup
#| include: false

# custom functions ----

library(ggplot2)

theme_standard <- theme_classic() +
  theme(
    axis.text = element_text(size = 11, color = "black"),
    axis.title = element_text(size = 16, color = "black"),
    axis.title.y = element_text(vjust = 1.5, color = "black"),
    axis.line = element_line(color = "black"),
    
    legend.position = "bottom",
    
    panel.background = element_rect(fill = "white", color = NA), 
    panel.border = element_blank(), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    strip.background = element_rect(fill = "grey95", color = "black"), 
    strip.text.x = element_text(size = 16, color = "black"),
    strip.text.y = element_text(size = 16, color = "black")) 

theme_facet <- theme_classic() +
  theme(
    axis.text = element_text(size = 11, color = "black"),
    axis.title = element_text(size = 16, color = "black"),
    axis.title.y = element_text(vjust = 1.5, color = "black"),
    axis.line = element_line(color = "black"),
    
    legend.position = "bottom",
    
    panel.background = element_rect(fill = "white", color = NA), 
    panel.border = element_rect(fill = NA, color = "black"), 
    panel.grid.major = element_line(color = "grey85"), 
    panel.grid.minor = element_blank(), 
    strip.background = element_rect(fill = "grey95", color = "black"), 
    strip.text.x = element_text(size = 16, color = "black"),
    strip.text.y = element_text(size = 16, color = "black")) 



# set options
knitr::opts_chunk$set(
  tidy = FALSE, 
  message = FALSE,
	warning = FALSE)

options(htmltools.dir.version = FALSE)

```


**Learning Objectives**

After completing this tutorial you should be able to

* describe detailed characteristics of a figure that is clear, accessible, transparent, and honest.
* describe characteristics of an effective figure legend.
* describe characteristics of an effective description of results.
* describe characteristics of an effective interpretation/discussion of central results.


[Download the directory for this project here](https://drive.google.com/drive/folders/1VNfSO6AA7eom-MKIxulW1D5Q-tbtZw2S?usp=sharing), make sure the directory is unzipped and move it to your `bi328` directory. You can open the `Rproj` for this module either by double clicking on it which will launch `Rstudio` or by opening `Rstudio` and then using `File > Open Project` or by clicking on the `Rproject` icon in the top right of your program window and selecting `Open Project`. 

There should be a file named `18_visualizations.qmd` in that project directory. Use that file to work through this tutorial - you will hand in your rendered ("knitted") `quarto` file as your homework assignment. So, first thing in the `YAML` header, change the author to your name. You will use this `quarto` document to record your answers. Remember to use comments to annotate your code; at minimum you should have one comment per code set^[You should do this whether you are adding code yourself or using code from our manual, even if it isn't commented in the manual... especially when the code is already included for you, add comments to describe how the function works/what it does as we introduce it during the participatory coding session so you can refer back to it.] you may of course add as many comments as you need to be able to recall what you did. Similarly, take notes in the document as we discuss discussion/reflection questions but make sure that you go back and clean them up for "public consumption".

Before you get started, let's make sure to read in all the packages we will need.

```{r}

# load libraries

# reporting
library(knitr)

# visualization
library(ggplot2)
library(ggthemes)
library(patchwork)

# data wrangling
library(dplyr)
library(tidyr)
library(readr)
library(skimr)
library(janitor)
library(magrittr)

# turn off scientific notation
options(scipen=999)

```


## Best practices for effective figures

Effective figures are used to tell a story. You can think of your exploratory analysis as the first step in determining what story the data tells (a fact finding mission if you will). Once you have determined what your specific question hypothesis is, the next step is narrowing down what analysis you will perform. Then you will identify your central results and interpret them in the context of your question. Your next step is communicating your results. At this point you need to identify which figures serve to communicate individual, specific points in the overall narrative.

In our example, we have refined our question to "Have impacts due to natural disasters increased at a global scale?". To answer this question, we are calculating the total annual loss of life and annual total economic damages incurred for each disaster type.

```{r}

disaster <- read_delim("data/nat-disasters_emdat-query-2021-10-05.txt", delim = "\t", skip = 6) %>%
  clean_names()

total_yr <- disaster %>%
  filter(!is.na(total_damages_000_us)) %>%
  group_by(year,disaster_type) %>%
  summarize(total_damages_yr = sum(total_damages_000_us))

deaths <- disaster %>%
  filter(!is.na(total_deaths)) %>%
  group_by(year, disaster_type) %>%
  summarize(total_deaths_yr = sum(total_deaths)) 

```

Then we created bubble plot showing the total loss of life and total damages incurred for each disaster type in each year.

```{r}
#| fig-width: 7
#| fig-height: 11
#| fig-cap: "Impacts due to natural disasters (1900 - 2020). Impacts were calculated as total annual economic costs (top) and total annual loss of life (bottom) due to each disaster type per year. The size of the bubble is scaled to cost [$] and total deaths [100k]. Data was obtained from EMDAT/International Disaster Database"
#| echo: false

p1 <- ggplot(total_yr, aes(x = year, y = disaster_type, size = total_damages_yr/1000000, fill = disaster_type)) +
  geom_point(shape = 21, alpha = .5) +
  scale_fill_viridis_d() +
  scale_size_continuous(breaks = c(5, 10, 25, 50, 100, 200),
                        name = "damages [USD]") +
  labs(x = "total damages incurred [USD]", y = " ") +
  theme_standard +
  theme(legend.position = "bottom") + 
  guides(fill = FALSE,
         size = guide_legend(override.aes = list(fill = "black", alpha = 1)))

p2 <- ggplot(deaths, aes(x = year, y = disaster_type, size = total_deaths_yr/100000, fill = disaster_type)) +
  geom_point(shape = 21, alpha = .5) +
  scale_fill_viridis_d() +
  scale_size_continuous(breaks = c(.5, 1, 5, 10, 15, 30),
                        name = "100k deaths") +
  labs(x = "total deaths incurred", y = " ",
       title = "Total deaths due to natural disasters (1900 - 2020)") +
  theme_standard +
  theme(legend.position = "bottom") + 
  guides(fill = FALSE,
         size = guide_legend(override.aes = list(fill = "black", alpha = 1)))

p1 / p2

```

::: {.callout-note icon=false}

## {{< fa clipboard-question >}}   Consider this

Summarize the overall narrative for the "story" we want to communicate based on our central question and state the specific point that each of these figures makes.

:::


::: {.callout-important icon=false}

## {{< fa pen >}}   Did it!

[Your answer here]

:::


::: {.callout-note icon=false}

## {{< fa clipboard-question >}}   Consider this

Data visualizations should be clear, accessible, transparent, and honest. For each of these categories list at least three criteria/guidelines that detail how data visualizations fulfill this category. Assess whether our figure(s) above fulfill these criteria and list at least three things that could be improved based this set of criteria.

:::


::: {.callout-important icon=false}

## {{< fa pen >}}   Did it!

[Your answer here]

:::


## Best practices for communicating results

Visualizations are a key component to effectively communicating your results. A good rule of thumb is that your written description of your results should allow the reader to understand what you are trying to communicate even without visualizations and that your title & legend should be descriptive enough that even without the written results (some would even say methods) the reader would understand the results as well.

::: {.callout-note icon=false}

## {{< fa clipboard-question >}}   Consider this

Give a description of what the components of an effective figure title & legend should look like and assess whether the figure legend above fulfills these criteria. List points for improvement.

:::


::: {.callout-important icon=false}

## {{< fa pen >}}   Did it!

[Your answer here]

:::


::: {.callout-note icon=false}

## {{< fa clipboard-question >}}   Consider this

Give a description of what an effective description of results should look like and write a description for the figure(s) above.

:::


::: {.callout-important icon=false}

## {{< fa pen >}}   Did it!

[Your answer here]

:::

::: {.callout-note icon=false}

## {{< fa clipboard-question >}}   Consider this

Give a description of the components of an effective discussion and write a brief discussion of the results of our example above (Refer back to the end of chapter 16 for this ...you can use your answer to the last question as a starting point).

:::


::: {.callout-important icon=false}

## {{< fa pen >}}   Did it!

[Your answer here]

:::



## Practice the best practices!

::: {.callout-note icon=false}

## {{< fa clipboard-question >}}   Consider this

Use one of the questions you explored for chapter 17 and complete the following tasks:

* State the overall broad question you are asking
* Summarize your "methods" in 2-3 sentences: How are you transforming/analyzing the data to specifically answer your question?
* Summarize the narrative of the story you want to communicate with this analysis and specify how the visualization serves the overall narrative.
* Generate your figure. Use bullet points to assess whether your visualization is clear, accessible, transparent, and honest.
* Write an effective title/legend for your figure.
* Write an effective description of your results.
* Write an effective interpretation discussion/interpretation of your results.

:::


::: {.callout-important icon=false}

## {{< fa pen >}}   Did it!

[Your answer here]

:::


## Spatial visualization

You have probably noticed that for some of the geographic comparisons were are making the most effective way to communicate those results would be to create a map with countries color coded according to the metric we are assessing.

Let's learn a straightforward method for visualization of spatial data. Run each of the code chunks below and comment the code line by line to describe what each function/argument is doing.

Install these packages as needed:

```{r}
#| eval: false

install.packages("rnaturalearth")

install.packages("rnaturalearthdata")

install.packages("sf")

```

Now we can load these libraries.

```{r}

library(rnaturalearth)
library(rnaturalearthdata)
library(sf)

```

The first thing we need is is some maps! The package `rnaturalearth` provides a map of countries of the entire world. Let's load a map as an `sf` class.

```{r}

world <- ne_countries(scale = "medium", returnclass = "sf")

class(world)

```

We see that world is both a `data.frame` and a class `sf`(simple features) which is a class of R object designed specifically for plotting maps.

Let's take a look at the column names

```{r}

colnames(world)

```

You can see that it contains quote a bit of information including some data about each country. The column `geometry` has the information about the country "shapes" (multipolygons) we want to plot.

```{r}

world %>%
  select(geometry) %>%
  head()

```

We can plot the map using `ggplot::geom_sf()` and the same syntax we generally use. We do not need to specify x or y coordinates as ggplot recognizes that we have passed a object of the class `sf` and that it contains a column called `geometry`.

```{r}

ggplot(data = world) +
  geom_sf()

```

We can manipulate the `geom_sf` layer using the same arguments we have used for other plot types. For example, let's make all the countries orange.

```{r}

ggplot(data = world) + 
    geom_sf(color = "black", fill = "orange") 

```

Or, we could color code them according to the column `pop_est` in the `world` data.frame which contains population estimates for each country.

```{r}

ggplot(data = world) +
    geom_sf(aes(fill = pop_est)) +
    scale_fill_viridis_c(trans = "sqrt") +
  coord_sf() 

```

Let's say we want to create a map that compares the number of droughts that occurred in each country from 2010 - 2020. First we would need to transform our raw data.

```{r}

droughts <- disaster %>%
  filter(year >= 2010 & year <= 2020) %>%
  group_by(iso) %>%
  summarize(droughts_total = n())

```

Now we can use `left_join()` to join the `world` and `drought` data.frames. Note that I used the column `iso` to calculate the number of droughts, this is equivalent to the column `iso_a3` in the `world` database. These are internationally recognized codes that designate every country and most independent areas with either a two or in this case three-letter abbreviations.

```{r}

world <- world %>%
  left_join(droughts, by = c("iso_a3" = "iso"))

head(world)

```

Now our `world` data.frame has a column specifying the number of droughts that occurred.

```{r}

ggplot(data = world) +
    geom_sf(aes(fill = droughts_total)) +
    scale_fill_viridis_c(trans = "sqrt") +
    coord_sf(expand = FALSE) +
  labs(x = "longitude", y = "latitude") +
  theme_bw() +
  theme(legend.position = "bottom")

```

::: {.callout-note icon=false}

## {{< fa clipboard-question >}}   Consider this

Use the figure we just generated and go through our steps for creating good visualizations according to our best practices:

* State the overall broad question you are asking
* Summarize your "methods" in 2-3 sentences: How are you transforming/analyzing the data to specifically answer your question?
* Summarize the narrative of the story you want to communicate with this analysis and specify how the visualization serves the overall narrative.
* Generate your figure - use the code above but optimize the figure as needed. Use bullet points to assess whether your visualization is clear, accessible, transparent, and honest.
* Write an effective title/legend for your figure.
* Write an effective description of your results.
* Write an effective interpretation discussion/interpretation of your results.

:::


::: {.callout-important icon=false}

## {{< fa pen >}}   Did it!

[Your answer here]

:::
