
# Making good figures

```{r}
#| label: setup
#| include: false

# custom functions ----

library(ggplot2)

theme_standard <- theme_classic() +
  theme(
    axis.text = element_text(size = 11, color = "black"),
    axis.title = element_text(size = 16, color = "black"),
    axis.title.y = element_text(vjust = 1.5, color = "black"),
    axis.line = element_line(color = "black"),
    
    legend.position = "bottom",
    
    panel.background = element_rect(fill = "white", color = NA), 
    panel.border = element_blank(), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    strip.background = element_rect(fill = "grey95", color = "black"), 
    strip.text.x = element_text(size = 16, color = "black"),
    strip.text.y = element_text(size = 16, color = "black")) 

theme_facet <- theme_classic() +
  theme(
    axis.text = element_text(size = 11, color = "black"),
    axis.title = element_text(size = 16, color = "black"),
    axis.title.y = element_text(vjust = 1.5, color = "black"),
    axis.line = element_line(color = "black"),
    
    legend.position = "bottom",
    
    panel.background = element_rect(fill = "white", color = NA), 
    panel.border = element_rect(fill = NA, color = "black"), 
    panel.grid.major = element_line(color = "grey85"), 
    panel.grid.minor = element_blank(), 
    strip.background = element_rect(fill = "grey95", color = "black"), 
    strip.text.x = element_text(size = 16, color = "black"),
    strip.text.y = element_text(size = 16, color = "black")) 



# set options
knitr::opts_chunk$set(
  tidy = FALSE, 
  message = FALSE,
	warning = FALSE)

options(htmltools.dir.version = FALSE)

```


**Learning Objectives**

After completing this tutorial you should be able to

* 


[Download the directory for this project here](), make sure the directory is unzipped and move it to your `bi328` directory. You can open the `Rproj` for this module either by double clicking on it which will launch `Rstudio` or by opening `Rstudio` and then using `File > Open Project` or by clicking on the `Rproject` icon in the top right of your program window and selecting `Open Project`. 

There should be a file named `18_figure-types.qmd` in that project directory. Use that file to work through this tutorial - you will hand in your rendered ("knitted") `quarto` file as your homework assignment. So, first thing in the `YAML` header, change the author to your name. You will use this `quarto` document to record your answers. Remember to use comments to annotate your code; at minimum you should have one comment per code set^[You should do this whether you are adding code yourself or using code from our manual, even if it isn't commented in the manual... especially when the code is already included for you, add comments to describe how the function works/what it does as we introduce it during the participatory coding session so you can refer back to it.] you may of course add as many comments as you need to be able to recall what you did. Similarly, take notes in the document as we discuss discussion/reflection questions but make sure that you go back and clean them up for "public consumption".

Before you get started, let's make sure to read in all the packages we will need.

```{r}

# load libraries

# reporting
library(knitr)

# visualization
library(ggplot2)
library(ggthemes)
library(patchwork)

# data wrangling
library(dplyr)
library(tidyr)
library(readr)
library(skimr)
library(janitor)
library(magrittr)

# turn off scientific notation
options(scipen=999)

```


## 
